---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Libraries and global options
```{r setup,results=F,message=F,warning=F}
library(tidyverse)
library(VIM)

#Set echo to True for all chunks, no messages and no warnings
knitr::opts_chunk$set(echo = T,
                      message = FALSE, 
                      warning = F)

```


## 1) Loading and preprocessing the data
```{r load}
activity<-read_csv("activity.csv")


```



## 2) What is mean total number of steps taken per day?

In the next table the total steps per day are shown

```{r steps per day,fig.width=3}

#Object with the sum of steps per day
act_per_day<-
      activity %>% 
      group_by(date)%>%
      summarise(steps_per_day=sum(steps,na.rm=T)) %>% 
      ungroup()

knitr::kable(act_per_day,format = "html", table.attr = "style='width:30%;'")
```

```{r mean and median}

#histogram
act_per_day %>% 
      ggplot(aes(steps_per_day))+
      geom_histogram(
            fill="dodgerblue4",binwidth = 2000
                  )

#mean and median
act_per_day %>% 
      summarise(mean=mean(steps_per_day,na.rm=T),
                median=median(steps_per_day,na.rm=T))

```

The mean is 9354 steps per day, and the median is 10395 steps.

## 3) What is the average daily activity pattern?

```{r daily pattern}

#mean per interval across al days
act_per_interval<-
   activity %>% 
   group_by(interval)%>%
   summarise(steps_per_interval=mean(steps,na.rm=T)) %>% 
   ungroup()

#time series plot
act_per_interval %>% 
   ggplot(aes(interval,steps_per_interval))+
   geom_line(color="#00BFC4",
             size=1)

#interval with the maximum number of steps:
act_per_interval$interval[act_per_interval$steps_per_interval==max(act_per_interval$steps_per_interval)]


```
The interval with the maximum number of steps is 835.

## 4) Imputing missing values


```{r missing values}
#TOTAL NUMBER OF ROWS
sum(is.na(activity$steps))

```
There are 2304 missing values in the variable steps.

Now I imput the missing values using KNN (k=5)
```{r imputation}
act_imp<-activity %>% 
   mutate(
      #I calculate the day of the week to use it as a variable for the imputation 
      day_of_the_week=weekdays(date)
   ) 

#KNN imputation (k=5)
#new dataset:
act_imp<-
   kNN(act_imp,
       imp_var = F
             )


```
New histogram, mean, and median with imputed data
```{r new mean and median}

#Object with the sum of steps per day
act_imp_per_day<-
   act_imp %>% 
   group_by(date)%>%
   summarise(steps_imp_per_day=sum(steps)) %>% 
   ungroup()

#histogram
act_imp_per_day %>% 
      ggplot(aes(steps_imp_per_day))+
      geom_histogram(
            fill="#F8766D",binwidth = 2000
                  )

#mean and median
act_imp_per_day%>% 
      summarise(mean=mean(steps_imp_per_day),
                median=median(steps_imp_per_day))



```
The mean has increased in 564 steps, but the median hasn't changed. 

Now I will analyze how the total steps per day have changed.

```{r total steps daily}

act_difference<-
   act_per_day %>% 
   left_join(act_imp_per_day) %>% 
   mutate(dif=steps_imp_per_day-steps_per_day)


act_difference %>%
  filter(dif!=0)
```

There are only differences between the original dataset and the imputed one in the days that all the values are missing values. That's the reason why the mean increases.



## 5) Are there differences in activity patterns between weekdays and weekends?

```{r weekdays and weekends}
act_imp<-act_imp %>% 
   mutate(
      type_of_day=factor(
         if_else(
            #my system is in Spanish. Sábado=Saturday, Domingo=Sunday
            day_of_the_week %in% c("sábado","domingo"),
            "weekend",
            "weekday"
         ),
         levels=c("weekend","weekday")
          )
      ) 

act_type_of_day<-act_imp %>% 
   group_by(type_of_day,interval) %>%
   summarise(steps_int_week=mean(steps)) %>% 
   ungroup()

act_type_of_day %>% 
   ggplot(aes(interval,steps_int_week,color=type_of_day))+
   geom_line(size=1)+
   facet_wrap(~type_of_day,nrow=2,ncol=1)+
   guides(color=F)+
   xlab("Interval")+
   ylab("Number of steps")+
   theme()

```

There are different patterns in activity between weekends and weekdays. At weekdays, between the intervals 540 and 905, the steps are significantly higher than weekends.In the rest of the intervals there are usually more steps at weekends.
